---
title: "DECDG: Data Analysis Exercise"
author: "Anton Prokopyev"
date: "September 4, 2017"
output:
  html_document:
    css: style.css
---
## How Life Expectancy Differs Between Wealthy and Developing Countries

Here at The World Bank, we are interested in using open data to share insights about economic development. In this report, we take a quick look at the how life expectancy differs between rich and poor countries, and display some of the new interactive data discovery tools we have been working on.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, results='hide', echo=FALSE, message=FALSE}
#### 0. Install and launch existing packages and API clients ####

# Specify the packages of interest
packages = c("WDI", "ggplot2", "dplyr", "ggpubr", "grid", "sqldf", "ggrepel", "plotly")

# Use this function to check if each package is on the local machine
# if a package is installed, it will be loaded
# if any are not, the missing package(s) will be installed and loaded
package.check <- lapply(packages, FUN = function(x) {
    if (!require(x, character.only = TRUE)) {
        install.packages(x, dependencies = TRUE)
        library(x, character.only = TRUE, quietly = TRUE)
    }
})
# Use this command if you need to verify which packages are loaded
# search() 
```

```{r, echo=FALSE, results='hide'}
# 1. Get the data and the metadata
library(WDI)
data_wdi <- WDI(indicator=c('SP.DYN.LE00.IN',
                            'SP.DYN.LE00.MA.IN',
                            'SP.DYN.LE00.FE.IN',
                            'SP.POP.TOTL'),
               country='all', 
               start=1960, 
               end=2015,
               extra=TRUE) # A parameter for metadata
```

```{r, echo=FALSE}
# 2. Prepare data for the analysis

# 2.1 Calculate the gender gap
data_wdi$gender_gap <- data_wdi$SP.DYN.LE00.FE.IN-data_wdi$SP.DYN.LE00.MA.IN
```

```{r, echo=FALSE, results='hide'}
# 2.2 Hadling data with 'sqldf' package to demonstrate some SQL skills

# Arrange original data in a shape ready for easy visualization
library(sqldf)
data_wdi_2015 <- sqldf('SELECT * 
                      FROM data_wdi
                      WHERE year = 2015 
                      AND region != "Aggregates" 
                      AND "SP.POP.TOTL" IS NOT NULL
                      AND gender_gap IS NOT NULL
                      AND income != "Not classified"
                     ')
```

#### Gender differences in life expectancy at birth
Using the 2015 WDI indicators as the most recent full snapshot of the world economy.
```{r, echo=FALSE, results='hide'}
# 3 Plotting

# 3.1 Set up some parameters for the figure

# Chose a stable y-axis scale
plt_ymin <- -0.5
plt_ymax <- 15

# Create uniform labels
plt_xlab <- "Years"
plt_ylab <- "Gender Gap, years"

# Define colorblind-frienly palettes 
cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", 
               "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
cbPalette2 <- c("#56B4E9", "#009E73", "#F0E442", "#0072B2", 
                "#D55E00", "#CC79A7", "#999999", "#E69F00")
cbPalette3 <- c("#F0E442", "#0072B2", "#D55E00", "#CC79A7", 
                "#999999", "#E69F00", "#56B4E9", "#009E73")
cbPalette4 <- c("#D55E00", "#CC79A7", "#999999", "#E69F00", 
                "#56B4E9", "#009E73", "#F0E442", "#0072B2")
cbPalette5 <- c("#0072B2", "#56B4E9", "#009E73", "#D55E00", 
                "#E69F00", "#CC79A7", "#F0E442", "#999999")
```

```{r, echo=FALSE}
# 3.2 Create the main figure

# Make sure text annotation and grid libraries are loaded along with ggplot2
library(ggplot2)
library(ggrepel) 
library(grid)

# Reorder 'levels' for 'income' variable in order to have a logical order in the legend 
data_wdi_2015$income <- factor(data_wdi_2015$income, 
                               levels = c("High income: OECD",
                                          "High income: nonOECD",
                                          "Upper middle income",
                                          "Lower middle income",
                                          "Low income",
                                          "Aggregates",
                                          "Not classified"))

# Proceed to plotting
p <- ggplot(data_wdi_2015,aes(SP.DYN.LE00.MA.IN, SP.DYN.LE00.FE.IN, SP.POP.TOTL, 
            size = SP.POP.TOTL, colour = factor(income))) +
          # Set graph type and the point transparency
            geom_point(alpha = 0.75) +
          # Adjust bubble sizes, remove a redundant legend
            scale_size(range = c(0.5, 12), guide = FALSE) +
          # Switch to square shapes for legend to avoid confusion
            guides(colour = guide_legend(override.aes = list(size=6, shape=15))) + 
          # Set legend position and font size
            theme(legend.position = 'right', legend.text=element_text(size=10)) + 
          # Select the black and white theme
            theme_bw() + 
          # Select a colorblind-friendly palette
          scale_colour_manual(values=cbPalette5) +
          # Remove legend title
            theme(legend.title=element_blank()) + 
          # X-axis label
            xlab("Male life expectancy at birth (years)")+
          # Y-axis label
            ylab("Female life expectancy at birth (years)") +
          # Make axis fonts smaller 
            theme(
            axis.title.x = element_text(size=9.5),
            axis.title.y = element_text(size=9.5)) +
          # Create automatic text captions for top-10 countries by population
            geom_text_repel( 
                    aes(label=ifelse(
                    SP.POP.TOTL>126000000, # Caption if population > 126M
                    as.character(country),'')),
                    size=3.22, color="black", force = 9, 
                    segment.color="#000000", segment.size = 0.25,
                    point.padding = unit(0.6, "lines"),
                    box.padding = unit(0.5, "lines"),
                    nudge_y = 1) +
          # Add a red dotted 'zero-gap' line
            geom_abline(intercept = 0, colour="red", alpha=0.5, linetype=3)
          
          # Print the plot
          p
          # Overlay the plot with metadata
          grid.text("Data Source: WDI, 2015 \nDate Created: 08/30/2017 \nAuthor: Anton Prokopyev, WBG",
          just = 0, x = 0.8, y = 0.059, # Choosing the position for metadata box
          gp = gpar(fontsize = 6, col = "black")) 
```

The red dotted line shows an ideal situation of gender parity, where males and females would have an equal life expectancy. Members of both genders would be expected to live for the same amount of time, e.g. 70 years for males (x-axis) and females (y-axis).

The point sizes vary to reflect differences in populations of these countries. 

We extracted several insights from the data and its visualization:

- Women do live longer than men everywhere except Swaziland. The country suffers from the lowest life expectancy in the world. The trend with female life expectancy dropping below the male indicator hit a low in 2004 and is potentially related to HIV situation in Swaziland. Explore the time trends further in this [line graph](https://data.worldbank.org/indicator/SP.DYN.LE00.FE.IN?locations=SZ). 

- Low and lower middle income countries are grouped in the lower left quadrant. This proves that people in developing economies live shorter than in rich countries. Interestingly, the gender gap in life expectancy in developing countries is smaller than that of wealthier countries. In very simple terms, if you are poor you will live a shorter life in any case. Being born female would not provide you with a major extension of the length of your life. Among others, this contrasts sharply with situation in Russia, an upper middle income country. The gender differences reach 11+ years there due to alcohol and tobacco policy shortcomings.

- It is unusual to see an upper middle income and a high income country among low income countries at the [55;60] mark. In this case, the two countries with abnormally low life expectancies are South Africa and Equatorial Guinea. The latter is a curious example of a resource curse. It is classified as a high income country due to high GNI per capita caused by the presence of oil resources (read more about groupings of economies [here](https://datahelpdesk.worldbank.org/knowledgebase/articles/378831-why-use-gni-per-capita-to-classify-economies-into)).

To explore these and other cases use zoom and panning in the interactive chart below.

```{r, echo=FALSE}
# 3.3 Create a clickable, interactive version of the same plot

# Load an interactive dataviz library
library(plotly)

# ! 
# It is recommended that you use the dev version of ggplot2 with plotly.
# Install it with: devtools::install_github('hadley/ggplot2')"

# Copy data for dataviz to keep the originals clean
data_wdi_2015_plotly <- data_wdi_2015

# Rename columns for dataviz purposes
names(data_wdi_2015_plotly)[names(data_wdi_2015_plotly) == 'SP.DYN.LE00.MA.IN'] <- 'Males'
names(data_wdi_2015_plotly)[names(data_wdi_2015_plotly) == 'SP.DYN.LE00.FE.IN'] <- 'Females'
names(data_wdi_2015_plotly)[names(data_wdi_2015_plotly) == 'SP.POP.TOTL'] <- 'Population'
names(data_wdi_2015_plotly)[names(data_wdi_2015_plotly) == 'income'] <- 'Group'
names(data_wdi_2015_plotly)[names(data_wdi_2015_plotly) == 'country'] <- 'Country'

p2 <- ggplot(data_wdi_2015_plotly,aes(`Males`, `Females`, 
            size = `Population`, colour = `Group`, label = `Country`)) +
              geom_point(alpha = 0.75) +
          # Adjust bubble sizes, remove a redundant legend
            scale_size(range = c(0.25, 12), guide = FALSE) +
          # Set legend position and font size
            theme(legend.position = 'bottom') + 
          # Select the black and white theme
            theme_bw() + 
          # Select a colorblind-friendly palette
          scale_colour_manual(values=cbPalette5) +
          # Remove legend title
            theme(legend.title=element_blank()) + 
          # X-axis label
            xlab("Male life expectancy at birth (years)") +
          # Y-axis label
            ylab("Female life expectancy at birth (years)") +
          # Add a no-gap line
            geom_abline(intercept = 0, colour="red", alpha=0.5, linetype=3)
          
          # Render the visualization with dynamic axis ticks
          ggplotly(p2, dynamicTicks=TRUE)  
```  
<br />  

This information provides important context for anyone working in economic development. Whether an economist or a policymaker, both can put it to use in the context of their work. Moving forward with this topic, we recommend augmenting the analysis by looking at life expectancy at age 65, similar to the analysis in this [paper](http://www.longevitypanel.co.uk/_files/life-expectancy-by-gender.pdf).

The World Bank staff are here to facilitate the access to development information. In future blog posts, we will be updating our dataviz catalog with new products. If you are interested in exploring development data and downloading it for further analysis, visit [World Bank Data](data.worldbank.org) to learn more.  

Reproducible research: download [RMarkdown file](https://prokopyev.github.io/wbg-projects/decdg_01/decdg_data_analysis_v1.Rmd).  
